---
title: 'Practical Machine Learning: Course Project'
author: "Luis P."
date: "05/19/2015"
output: html_document
---

<2000 words, <=5 figures

http://lprisan.github.io/PracML_PeerAssessment

Do the 20 predictions of the test set (auto grading)


# Executive Summary

Goal: predict the manner in which they exercise (classe variable)


# Loading and preprocessing of data

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
(data by http://groupware.les.inf.puc-rio.br/har)

Maybe move this to an appendix? or just not show

```{r}
require(caret)

download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "./data/dataTrain.csv", method="curl")
train <- read.csv("./data/dataTrain.csv")
summary(train)
```

... eliminate the variables that are not going to be tested, to simplify the model?

... exploratory analysis (do variables look gaussian?) good if we want to do PCA

# Building the model


## Cross-validation

## Expected out-of-sample error

## Rationale


# Testing the model

In-sample error: Accuracy % missed, specificity, sensitivity, +/- predictive value, ROC curves?

OOS/Testing error: Accuracy % missed, specificity, sensitivity, +/- predictive value, ROC curves?

```{r}
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "./data/dataTest.csv", method="curl")
test <- read.csv("./data/dataTest.csv")
summary(test)
```